\hydrogen{} and \carbon{} Network Strategies for the European Energy System

### 2024-09-06 report/submissions/journal/paper/paper.tex
update scenarios, continue with review comments

We analyze how the deployment of carbon and hydrogen networks affects different decarbonization strategies in the scenarios and how investments in none or only one of the transport technologies lead to inefficient market layouts. We do this by focusing first on a net-zero \carbon{} emissions target while limiting carbon sequestration to 200~Mt/a. We give a detailed description of the different technological and geographical impacts of the transportation systems. Subsequently, our analysis demonstrates the extent to which different deployment strategies remain robust under a tightened net \carbon{} removal target.

@@

All scenarios share key energy system characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). Costs in the \baselinescenario{} are the highest at 764~bn€/a. Scenarios with carbon and hydrogen networks are more cost-effective, reducing costs by 3.1\% in the \carbongrid{}, 4.7\% in the \hydrogengrid{}, and 5.3\% in the \hybridscenario{}.

@@

% In every model, nearly half the costs are attributable to wind and solar electricity production, and about 10\% to hydro, nuclear, and biomass. In the \baselinemodel, transmission costs (electricity, gas, hydrogen, carbon) are below 4\% (25 bn€) and are dominated by electricity grid expenses. In contrast, the other models have higher combined pipeline costs, peaking at 20\% in the \hybridmodel. Both carbonaceous fuel production and electrolysis account for about 4\% (30 bn€) and 4.7\% (37 bn€) of costs respectively, with heat pump installations contributing around 8\% (64 bn€/a in the \baselinemodel).

@@

\subsection*{Higher sequestration limits reintroduce fossil fuels}
\label{sec:higher_sequestration}
Carbon management and hard-to-abate sectors of the optimized system are strongly driven by the model's limited sequestration potential of 200~Mt/a. In the Hybrid scenario, this limitations imposes a shadow-price of 225€/t to the system, meaning the system would gain 225€ from an additional tonne of sequestration potential. As soon as the sequestration limit is increased, the model starts to reintroduce fossil fuels as a primary energy source in the the aviation and industrial sectors.

\begin{figure}[htb!]
    \centering
    \includegraphics[width=\linewidth]{comparison/sequestration/figures/90_nodes/cost_bar.png}
    \caption[short]{Total system cost as a function of the sequestration limit.}
    \label{fig:sequestration_cost_bar}
\end{figure}

@@

Limitations of the study, such as the neglect of synthetic fuel imports and assumptions on biomass potentials, are discussed in the Appendix in Section~\ref{sec:limitations}. A sensitivity analysis of the costs of \carbon{} pipelines, which show negligible effects on the results, can be found in the Appendix in Section~\ref{sec:subsidy}.

@@

To store carbon, we differentiate between short-term storage in steel tanks without permanent containment and long-term, irreversible sequestration in underground sequestration sites such as porous rock formations or depleted gas reservoirs.
Costs for both options are included in the model.
For carbon sequestration, we only consider offshore sites as potential sinks (see Fig.~\ref{fig:sequestration_map}).
We make this choice because offshore sequestration sites typically have a larger capacity compared to onshore sequestration sites in saline aquifers and due to concerns over public safety for infrastructure near populated areas.
Our estimates for carbon sequestration potentials are restrictive, limiting the annual sequestration potential to 25~Mt per region and constraining it further such that this amount could be  25 years.
Furthermore, we decided to limit the total amount of sequestered \carbon{} to 200~Mt per year for our net-zero \carbon{} emission target and 660~Mt per year for our net-negative emission target.
This is enough to offset hard-to-abate fossil process emissions (e.g., calcination of limestone in cement manufacturing) and some limited slack for fossil fuels to be used where they can be most valuable.
This is in order to avoid offsetting fossil emissions with carbon removal if they can technologically be avoided in the first place.

@@

\subsection{Subsidizing \carbon{} pipelines}
\label{sec:subsidy}

The effect of a 50\% subsidy on \carbon{} pipelines impacts the optimal technology deployment to a small extent. Fig.~\ref{fig:cost_bar_subsidy} shows the corresponding total system cost of all scenarios with a net carbon neutrality target. Fig.~\ref{fig:cost_bar_diff_subsidy} shows the net change in system cost between non-subsidized and subsidized models, showing only the \carbongrid{} and \hybridscenario{} where changes occur.

@@

\clearpage
\subsection{Cost comparison across scenarios}
\label{sec:cost_comparison}

>>>>

We analyze how the deployment of carbon and hydrogen networks affects different decarbonization strategies in the scenarios and how investments in none or only one of the transport technologies lead to inefficient market layouts. Initially, we focus on a net-zero carbon emissions target, limiting carbon sequestration to 200~Mt/a. This constraint ensures that carbon removal is reserved for the hardest-to-abate emissions, while avoiding reliance on sequestration where other mitigation options are feasible. We give a detailed description of the different technological and geographical impacts of the transportation systems. Subsequently, our analysis demonstrates the extent to which different deployment strategies remain robust under a tightened net \carbon{} removal target.

@@

All scenarios share key energy system characteristics but differ in system costs and technology use (Fig.~\ref{fig:cost_bar}). Costs in the \baselinescenario{} are the highest at 764~bn€a. Scenarios with carbon and hydrogen networks are more cost-effective, reducing costs by 3.1\% in the \carbongrid{}, 4.7\% in the \hydrogengrid{}, and 5.3\% in the \hybridscenario{}.

@@

% In every model, nearly half the costs are attributable to wind and solar electricity production, and about 10\% to hydro, nuclear, and biomass. In the \baselinemodel, transmission costs (electricity, gas, hydrogen, carbon) are below 4\% (25 bn€) and are dominated by electricity grid expenses. In contrast, the other models have higher combined pipeline costs, peaking at 20\% in the \hybridscenario. Both carbonaceous fuel production and electrolysis account for about 4\% (30 bn€) and 4.7\% (37 bn€) of costs respectively, with heat pump installations contributing around 8\% (64 bn€/a in the \baselinemodel).

@@

@@

Limitations of the study, such as the neglect of synthetic fuel imports and assumptions on biomass potentials, are discussed in the Appendix in Section~\ref{sec:limitations}.
The Appendix in Section~\ref{sec:sensitivities} provides an analysis of the model's sensitivities. Notably, relaxing the sequestration limit has a significant impact, reducing system costs by 9.1\% at 800~Mt/a in the hybrid scenario. This reduction is achieved by opting for fossil fuels with carbon capture, transport, and sequestration instead of carbon utilization. Therefore, investments in renewable power assets which serve the synthetic fuel production for the aviation, shipping and industry sectors drop by roughly a third. Less effects can be seen from reducing capital costs for \carbon{} pipelines and costs of electrolyzers, on one hand indicate robustness against price changes but also little effect of subsidizing these technologies.  As the cost optimum is affected by the sequestration potential limits, a more detailed analysis of a relaxed sequestration limit is presented in the Appendix in Section~\ref{sec:higher_sequestration}.

@@

To store carbon, we differentiate between short-term storage in steel tanks without permanent containment and long-term, irreversible sequestration in underground sequestration sites such as porous rock formations or depleted gas reservoirs.
Costs for both options are included in the model.
For carbon sequestration, we only consider offshore sites as potential sinks (see Fig.~\ref{fig:sequestration_map}).
We make this choice because offshore sequestration sites typically have a larger capacity compared to onshore sequestration sites in saline aquifers and due to concerns over public safety for infrastructure near populated areas.
Our estimates for carbon sequestration potentials are restrictive, limiting the annual sequestration potential to 25~Mt per region and constraining it further such that this amount could be  25 years.
Furthermore, we decided to limit the total amount of sequestered \carbon{} to 200~Mt per year for our net-zero \carbon{} emission target and 660~Mt per year for our net-negative emission target.
This constraint is deliberately implemented to avoid over-reliance on carbon capture and sequestration as a backstop technology, which could otherwise offset fossil emissions instead of prioritizing the reduction of fossil fuel use where technologically feasible. By capping the sequestration at 200 Mt/year, we aim to promote a balanced energy transition, where sequestration is used to offset hard-to-abate fossil process emissions (e.g., calcination of limestone in cement manufacturing) but does not become a substitute for more sustainable mitigation measures. The implications of this choice are discussed in the Appendix in Section~\ref{sec:higher_sequestration}.

This is enough to offset hard-to-abate fossil process emissions (e.g., calcination of limestone in cement manufacturing) and some limited slack for fossil fuels to be used where they can be most valuable.
This is in order to avoid offsetting fossil emissions with carbon removal if they can technologically be avoided in the first place.

@@

\clearpage
\subsection{Sensitivities}
\label{sec:sensitivities}

\subsubsection*{Higher sequestration limits}
\label{sec:higher_sequestration}
Carbon management in the optimized energy system, particularly in the hard-to-abate sectors, is heavily influenced by the model's sequestration potential, which is capped at 200~Mt/year. Lifting the sequestration limit leads to cost benefits and a shift in primary energy supply and its implications (see Fig.~\ref{fig:sequestration_cost_bar}). With higher sequestration rates, more fossil fuels are used in the aviation, heating, and industry sectors in land-locked regions where hydrogen is expensive. This reduces the reliance on carbon utilization (CU) and, therefore, the production and transport of hydrogen. At the same time, the carbon network is expanded to transport more carbon from inland point sources to sequestration sites. Sequestration is further supported by large DAC facilities near coastal areas to offset additional emissions from fossil fuels.
These effects lead to significant changes in investments when increasing sequestration from 200~Mt/a to 800~Mt/a. Due to reduced hydrogen production, investments in wind and solar drop by a third. Total system costs are reduced by 9.1\%, primarily due to lower renewable energy investments. Regarding transport networks, hydrogen network capacity is halved, while the carbon network quadruples (see Fig.~\ref{fig:sequestration_cost_bar_transmission}). These findings are consistent with the effects discussed in~\cite{hofmannDesigningCO2Network2023}.

\begin{figure}[ht!]
    \centering
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\linewidth]{comparison/sequestration/figures/90_nodes/cost_bar.png}
        \caption[short]{}
        \label{fig:sequestration_cost_bar}
    \end{subfigure}%
    \begin{subfigure}{.5\textwidth}
        \includegraphics[width=\linewidth]{comparison/sequestration/figures/90_nodes/cost_bar_transmission.png}
        \caption[short]{}
        \label{fig:sequestration_cost_bar_transmission}
    \end{subfigure}
    \caption{System costs for the \hybridscenario{} as a function of the sequestration limit (a). The transmission system cost for the \hybridscenario{} as a function of the sequestration limit (b).}
\end{figure}

As sequestration capacity increases, the cost savings from additional sequestration begin to plateau. At 200~Mt/a, the shadow price of the sequestration constraint shows that each additional tonne saves 225~€/t, but at 800~Mt/a, this saving drops to 51~€/t. Extrapolating this behavior, saturation of sequestration can be expected around 1200~Mt/a.

While increasing sequestration capacity offers economic benefits, it poses potential risks. Over-reliance on sequestration may delay necessary investments in cleaner technologies and renewable energy. Additionally, the large-scale expansion of carbon storage infrastructure may encounter public opposition due to concerns about the long-term stability of CO$_2$ storage sites and possible leakage risks. Furthermore, the focus on sequestration could perpetuate the use of fossil fuels in sectors where decarbonization alternatives, such as electrification or green hydrogen, may offer more sustainable solutions. Thus, careful consideration must be given to avoid locking in fossil fuel usage and undermining the broader energy transition goals.

\subsubsection{Lower \carbon{} pipelines costs}
\label{sec:subsidy}

The effect of a 50\% cost-reduction on \carbon{} pipelines, such as through subsidies, impacts the optimal technology deployment to a small extent. Fig.~\ref{fig:cost_bar_subsidy} shows the corresponding total system cost of all scenarios with a net carbon neutrality target. Fig.~\ref{fig:cost_bar_diff_subsidy} shows the net change in system cost between non-subsidized and subsidized models, showing only the \carbongrid{} and \hybridscenario{} where changes occur.

@@

\clearpage
\subsection{Other sectors in the energy system}

The following maps show the optimal supply, demand, transport and prices of energy carriers in the \hybridscenario{} not covered in the main text.

\begin{figure}[ht!]
    \centering
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_electricity.png}
        \caption{Optimal supply, consumption, transport and prices of electricity in the \hybridscenario{}.}
        \label{fig:balance_map_electricity}
    \end{subfigure}%
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_heat.png}
        \caption{Optimal supply, consumption, transport and prices of heat in the \hybridscenario{}.}
        \label{fig:balance_map_heat}
    \end{subfigure}
\end{figure}

\begin{figure}[ht!]
    \centering
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_gas.png}
        \caption{Optimal supply, consumption, transport and prices of gas in the \hybridscenario{}.}
        \label{fig:balance_map_gas}
    \end{subfigure}%
    \begin{subfigure}{0.5\textwidth}
        \includegraphics[width=\linewidth]{full/figures/90_nodes/balance_map_biogas.png}
        \caption{Optimal supply, consumption, transport and prices of biogas in the \hybridscenario{}.}
        \label{fig:balance_map_biogas}
    \end{subfigure}
\end{figure}

\clearpage
\subsection{Cost comparison across net-zero scenarios}
\label{sec:cost_comparison}
